---
title: "Analysis"
execute: 
  echo: false
format: html
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidycensus)
library(GGally)
library(ggplot2)
library(ggthemes)
library(tidymodels)
library(car)
library(marginaleffects)
library(broom)
```

## 1. Explanatory Data Analysis
Following is EDA.
```{r}
#| warning: false
multi_vars <- c(
  gini = "B19083_001",
  median_income = "B19013_001",
  poverty = "B17001_002",                    # Below poverty
  education_total = "B15003_001",           # Total population 25+
  education_bachelor = "B15003_022",        # Bachelor's degree
  labor_force = "B23025_003",               # Labor force
  unemployed = "B23025_005")                # Unemployed

socio_data <- get_acs(
  geography = "county",
  variables = multi_vars,
  year = 2020,
  survey = "acs5",
  output = "wide")

# Clean and calculate additional metrics
socio_clean <- socio_data |>
  transmute(
    county = NAME,
    gini = giniE,
    median_income = median_incomeE,
    poverty = povertyE,
    education_rate = education_bachelorE / education_totalE,
    unemployment_rate = unemployedE / labor_forceE)

```

#  1.1 Summary of Variables 
```{r}
summary(socio_clean)
```

#  1.2 Plot
```{r}
#| warning: false
plot_data <- socio_clean |>
  mutate(
    log_income = log(median_income),
    log_poverty = log(poverty + 1)) |>
  select(
    Gini = gini,
    `Log Income` = log_income,
    `Log Poverty` = log_poverty,
    `Education Rate` = education_rate,
    `Unemployment Rate` = unemployment_rate)

# Plot
ggpairs(
  plot_data,
  lower = list(continuous = function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) +
      geom_point(alpha = 0.6, color = "skyblue") +
      geom_smooth(method = "loess", se = FALSE, color = "red", linetype = "dashed", size = 0.5)}),
  upper = list(continuous = wrap("cor", size = 3)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.4, fill = "#69b3a2"))) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray30"),
    plot.caption = element_text(size = 10, color = "gray40"),
    strip.text.y = element_text(angle = 0, hjust = 1))  +
  labs(
    title = "Income Inequality with other Indicators in U.S",
    subtitle = "Education and Income have negative But Povery and Unemployment have positive relationship with Income Inequality",
    caption = "Source: American Community Survey (ACS) 2020")
```

<div style="text-align: justify;">Above plot shows that increse in Education and Income has reduced income inequality whereas, increse in unemployment and poverty has incresed income inequality. The relationship among income inequality (gini) and other social indicators can be seen in plot. Log of income and poverty were taken for better comparision purpose. Further, Histograms of every indicator can also be seen in dignal which shows that distributions of gini, log income, log poverty are normal but distributions of education rate and unemployment rate are slightly and significantly right skewed respectively.</div>

# 2. Data Normality 

<div style="text-align: justify;">Skewness in distribution can influnce the model, therfore, all the variables were scaled to have better results for comparision or explanation. Following code was used for log and scale.</div>

```{r} 
socio_clean <- socio_clean |>
  mutate(
    log_income = log(median_income),
    log_poverty = log(poverty + 1))  # add 1 to avoid log(0)
  
socio_scaled <- socio_clean |>
  mutate(across(c(log_income, log_poverty, education_rate, unemployment_rate),  scale))

socio_scaled_data <- socio_scaled |>
  select(gini, log_income, log_poverty, education_rate, unemployment_rate)
```

#  2.1 Histogram
```{r}
#| warning: false
plot_socio_scaled <- socio_scaled_data |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Plot histograms for all variables with faceting and custom theming
ggplot(plot_socio_scaled, aes(x = value, fill = variable)) +
  geom_histogram(bins = 40, color = "white", alpha = 0.85) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 14, color = "navy"),
    axis.title = element_blank(),
    axis.text = element_text(color = "gray30"),
    panel.grid.major = element_line(color = "gray85"),
    legend.position = "none",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 10, color = "gray40", hjust = 1)
  ) +
  labs(
    title = "Distribution of Variables After Scale",
    caption = "Source: American Community Survey (ACS) 2020"
  )
```

<div style="text-align: justify;">Above histograms show that distributions of gini, log_income, log_poverty are normal howerver, distributions of education_rate and unemploment_rate are still rightly skewed. Hence, scale of variables is purposeless.</div>

#  2.2 Shapiro test
<div style="text-align: justify;">Shapiro test also checks whether distribution is normal or not. If results show that W value is equal to 1 and p value is > 0.05 then distribution is normal.</div>

#  2.2.1 Unemployment Rate 
```{r}
# Data Normality test
unemployment_rate <- socio_scaled_data |> pull(unemployment_rate)
shapiro.test(unemployment_rate)
```

#  2.2.2 Education Rate
```{r}
education_rate <- socio_scaled_data |> pull(education_rate)
shapiro.test(education_rate)
```

<div style="text-align: justify;">In case of both variables; unemployment_rate and education_rate, shapiro showed that distributions are not normal.</div>

# 3.  Model
<div style="text-align: justify;">My outcome variable was continues therfore, Liner Regression Model was applied. Following is the general form of a linear regression model.</div>

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_p X_{pi} + \varepsilon_i
$$

Where:
- \( Y_i \) is the dependent (response) variable,
- \( X_{1i}, X_{2i}, \ldots, X_{pi} \) are the independent variables (predictors),
- \( \beta_0 \) is the intercept,
- \( \beta_1, \ldots, \beta_p \) are regression coefficients,
- \( \varepsilon_i \sim \mathcal{N}(0, \sigma^2) \) is the error term assumed to follow a normal distribution.

Probability Family Function when the outcome variable is normally distributed:

$$
Y_i \sim \mathcal{N}(\mu_i, \sigma^2), \quad \text{where} \quad \mu_i = \beta_0 + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}
$$

The likelihood function for all \( n \) observations is:

$$
L(\boldsymbol{\beta}, \sigma^2) = \prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( -\frac{(Y_i - \mu_i)^2}{2\sigma^2} \right)
$$


# 3.1 Coefficients

```{r}
socio_clean <- socio_clean |>
  mutate(
    income = median_income / 100000,
    poverty = poverty / 100000) |>
  select(gini, income ,poverty, education_rate, unemployment_rate )
```

```{r}
income_inequality1 <- linear_reg() |> 
   fit(gini ~ income + poverty + education_rate + unemployment_rate, data = socio_clean)
```

```{r}
income_inequality1 |> 
  tidy(conf.int = TRUE)
```

<div style="text-align: justify;">EDA show that education rate has negative correlation with income inequality but coeffients show that it predicts positively. Hence, need for Variance Inflation Factor (VIF) analysis to see whether the change of coefficient sign is due to multicollinearity.</div>

# VIF analysis

```{r}
vif(lm(gini ~ income + poverty + education_rate + unemployment_rate, data = socio_clean))
```

<div style="text-align: justify;">VIF between 1 to 5 means acceptable multicollinearity among predictors. However, beyound 5 means problematic. Here, VIF results show that there is no multicollinearity. Therefore, it was decided to remove education rate because it may disturbe the model results because in model coefficients its results sign is not consistant with EDA</div>

# 3.2 Coefficients without education
```{r}
fit_income_inequality <- linear_reg() |> 
   fit(gini ~ income + poverty + unemployment_rate, data = socio_clean)
```

```{r}
fit_income_inequality |> 
  tidy(conf.int = TRUE)
```

# 3.3 Data Generating Mechanism
$$
\hat{\text{Gini}} = 0.475 
- 0.0831 \cdot \text{Income} 
+ 0.0177 \cdot \text{Poverty}  
+ 0.265 \cdot \text{UnemploymentRate}
$$

# 3.4 Model Prediction

```{r}
fit_income_inequality |>
  tidy() |>
  filter(term %in% c("income", "poverty", "unemployment_rate")) |>
  ggplot(aes(x = term, y = estimate, color = term)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * std.error, 
        ymax = estimate + 1.96 * std.error),
    width = 0.3, 
    size = 1.2
  ) +
  scale_color_viridis_d() +
  labs(
    title = "Income Inequality Predicted by Income, Poverty, & Unemployment",
    subtitle = "Income predicts negatively, while poverty and unemployment predict positively",
    x = "Predictor Variable",
    y = "Predicted Estimate",
    caption = "Data source: ACS 2020"  # <- Fixed: Added missing quotation mark
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "gray95"),
    plot.background = element_rect(fill = "white", color = "white"),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_line(color = "white"),
    plot.title = element_text(face = "bold", size = 15.5, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 1),
    legend.position = "none"
  )
```

<div style="text-align: justify;"> Increase in income reduces income inequality by -0.083 on the other hand, increse in poverty and unemployment rate increse income inequality by 0.018 and 0.265 respectively. Unemployment rate predicts higher than others.</div>
