---
title: "Model"
author: Naveed Ahmad
format: html
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidycensus)
library(tidymodels)
# libraries needed for loading data and model
```

```{r}
#| warning: false
multi_vars <- c(
  gini = "B19083_001",                
  median_income = "B19013_001",
  poverty = "B17001_002",                    # Below poverty
  education_total = "B15003_001",           # Total population 25+
  education_bachelor = "B15003_022",        # Bachelor's degree
  labor_force = "B23025_003",               # Labor force
  unemployed = "B23025_005")                # Unemployed

socio_data <- get_acs(
  geography = "county",
  variables = multi_vars,
  year = 2020,
  survey = "acs5",
  output = "wide")
# Clean and calculate additional metrics
socio_clean <- socio_data |>
  transmute(
    county = NAME,
    gini = giniE,
    median_income = median_incomeE,
    poverty = povertyE,
    education_rate = education_bachelorE / education_totalE,
    unemployment_rate = unemployedE / labor_forceE)
```

```{r}
socio_clean1 <- socio_clean |>
  mutate(
    income = median_income / 100000,
    poverty = poverty / 100000) |>
  select(gini, income ,poverty, unemployment_rate )
```

## Model
<div style="text-align: justify;">My outcome variable was continues therfore, Liner Regression Model was applied. Following is the general form of a linear regression model.</div>

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_p X_{pi} + \varepsilon_i
$$

Where:
- \( Y_i \) is the dependent (response) variable,
- \( X_{1i}, X_{2i}, \ldots, X_{pi} \) are the independent variables (predictors),
- \( \beta_0 \) is the intercept,
- \( \beta_1, \ldots, \beta_p \) are regression coefficients,
- \( \varepsilon_i \sim \mathcal{N}(0, \sigma^2) \) is the error term assumed to follow a normal distribution.

Probability Family Function when the outcome variable is normally distributed:

$$
Y_i \sim \mathcal{N}(\mu_i, \sigma^2), \quad \text{where} \quad \mu_i = \beta_0 + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}
$$

The likelihood function for all \( n \) observations is:

$$
L(\boldsymbol{\beta}, \sigma^2) = \prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( -\frac{(Y_i - \mu_i)^2}{2\sigma^2} \right)
$$


# 3.2 Coefficients
```{r}
income_inequality <- linear_reg() |> 
   fit(gini ~ income + poverty + unemployment_rate, data = socio_clean1)
```

```{r}
income_inequality |> 
  tidy(conf.int = TRUE)
```

# 3.3 Data Generating Mechanism
$$
\hat{\text{Gini}} = 0.475 
- 0.0831 \cdot \text{Income} 
+ 0.0177 \cdot \text{Poverty}  
+ 0.265 \cdot \text{UnemploymentRate}
$$

